Swimming World Records
=======================

```{r include=FALSE}
require(mosaic)
require(fastR)
trellis.par.set(theme=col.mosaic())
```

The data set `SwimRecords` contains world records in the 100m freestyle race.

Men and women have different records:
```{r}
mean( time ~ sex, data=SwimRecords)
```

Of course, the record varies over the years:
```{r}
xyplot(time ~ year, data=SwimRecords )
```

Here is a template for fitting a model and overplotting the function on a graph.  You'll modify this in various ways.


```{r}
xyplot(time ~ year, data=SwimRecords )
mod0 = lm( time ~ 1+sex*year, data=SwimRecords)
f0 = makeFun(mod0)
plotFun(f0(sex="M",year=y)~y, add=TRUE)
plotFun(f0(sex="F",year=y)~y, col="red",add=TRUE)
```


```{r}
xyplot(time ~ year, data=SwimRecords )
mod1 = lm( time ~ year+sex, data=SwimRecords)
f1 = makeFun(mod1)
plotFun(f1(year=y,sex="M")~y, add=TRUE)
plotFun(f1(year=y,sex="F")~y, col="red",add=TRUE)
```

## Inference

### Confidence intervals
```{r}
s = do(100)*lm(time ~ year+sex, data=resample(SwimRecords))
densityplot(~year, data=s)
```

Or ...
```{r}
mod = lm(time~year+sex, data=SwimRecords)
confint(  )
```

Or ...
```{r}
summary(mod)
```

#### Hypothesis Testing

```{r}
s = do(100)*lm(time ~ shuffle(year)+sex, data=resample(SwimRecords))
densityplot(~year, data=s)
```
How many of the coefficients are bigger than those observed?


See the regression report for classical technique

#### ANOVA
```{r}
anova(mod)
```
## Wage Records

You must do the following in the console --- just hit "run"
```{r eval=FALSE}
fetchData("mLM.R") # do only once
mLM(wage ~ age*sex*educ + sector*sex, data=CPS85)
```